\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex

\vspace{-1em}
\begin{abstract}
Recombination is an ongoing and increasingly important feature of circulating
lineages of SARS-CoV-2, challenging how we represent the evolutionary history
of this virus and giving rise to new variants of potential public health
concern by combining transmission and immune evasion properties of different
lineages. Detection of new recombinant strains is challenging, with most
methods looking for breaks between sets of mutations that characterize distinct
lineages.
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real time and to cohesively integrate them into a
genealogy in the form of an ancestral recombination graph (ARG),
which jointly records mutation, recombination and genetic
inheritance. We infer two ARGs under
different sampling strategies, and study their properties.
One contains 1.27 million sequences
sampled up to June 30, 2021, and the second is more sparsely sampled,
consisting of 657K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and feature-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
genetic variation, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to utilise an efficient and convenient platform for computation
based on well-engineered technologies
makes \texttt{sc2ts} a promising approach.
\end{abstract}

@@

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from Nextstrain.
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome,
although other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the Nextstrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

The current version of \texttt{sc2ts} infers a
large number of ``reversion'' mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation (see Section~\ref{sec:args} for precise definitions).
These are symptomatic of both data quality issues
such as ``time travellers'' (Section~\ref{sec:filtering_time_travellers})
and problematic sites (e.g., 28,271 as discussed in
Section~\ref{sec:breakpoint_intervals}; see also
Figure~\ref{fig:pango_XB_gisaid_graph} for an example of multiple
reversions at this site), as well as indicating opportunities
for improvement in tree building hueristics
(Section~\ref{sec:parsimony-heuristics}).
For example, the current ``reversion push'' operation
only eliminates reversions in the newly added portion of the tree
and not reversions around earlier nodes created by this algorithm.

@@

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. Using Nextclade designations, 15 Pango X lineages are
inferred to have multiple
recombinant origins, such that their samples trace to more than one most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e., there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend and which agrees with the offical Pango designated parent lineages
(see Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

@@

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps,
is a signature feature of
SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterisation of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will be a poor match to
the existing ARG, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
segments.
This appears to be the case for the Delta VoC,
corresponding to rows A and B of Table~\ref{tab:false_positive},
which emerged and quickly rose to high prevalence in early-mid 2021,
carrying 30 characteristic mutations compared to the reference sequence
\citep{McCrone2022-context}, including nine mutations in the S gene
not seen in earlier VoCs.
The origins of the  Delta VoC are
are illustrated in  Figure~\ref{fig:false_positives},
which shows the large numbers of mutations involved,
and the multiple closely related recombinations inferred
before tree-like behaviour is resumed at the ancestor of
[FIXME what percentage?] Delta samples (node \texttt{tsk261771}).
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

@@

% PARA 2. Here are things that could be improved in the short term. Here's a
% few more, longer term ideas.
Nonetheless, \texttt{sc2ts} is currently ``alpha'' quality software, and
we caution against overinterpreting current results. As we have sought
to illustrate throughout, there are some clear areas for improvement.
The pipeline used to identify and
mask erroneous sites in the input alignments is simplistic, and, among
other issues, results in multi-base indels being marked as missing data
(Section~\ref{sec:data_preprocessing}). A more sophisticated
approach~\citep[e.g.,][]{Aksamentov2021-hj} would
likely yield significant improvements and reduce the effect of sites with
artefactually high levels of mutation (e.g., site 28271;
Section~\ref{sec:breakpoint_intervals}).
Using a pre-existing tree built using state-of-the-art phylogenetic methods
for the early stages of the pandemic (Section~\ref{sec:backbone_phylogeny}),
and minor adaptations to standard node-dating methods
(Section~\ref{sec:node_dating}) should help resolve the most notable issues with
the inferred backbone phylogeny (Figure~\ref{fig:cophylogeny}).
Trees constructed from daily sample clusters have a surprisingly large
influence on the overall ARG topology
(Section~\ref{sec:sample-cluster-tree-inference}), and so using a more
sophisticated tree building approach should yield clear improvements.
The unrealistically large number of reversion mutations
(Section~\ref{sec:mutation_spectrum}) may be reduced by
improvements to the current parsimony
heuristics (Section~\ref{sec:parsimony-heuristics}).
A major source of errors are the ``time-traveller'' samples, whose
recorded collection dates are months (or years) too early
(Section~\ref{sec:filtering_time_travellers}).
% Hmm, not sure about this. Wouldn't this also rule out new major lineages
% also (which tend to show up with a bunch of new mutations?) I guess it
% depends on the threshold, and we could set it according to previous
% experience with new lineages?
While it is unclear how
we might solve this problem in general, some simple solutions such
as filtering out sequences that exceed a given cost in the LS HMM
(i.e., number of mutations and recombination switches) may work well
in practice. Such an approach would also reduce the impact of sequences
with high levels of sequencing error (which currently contribute a large number of
mutations).
Taken together, these and other relatively minor improvements should
enable inference over much larger subsets of the dataset and
give a clearer picture of the combined processes of recombination and
mutation over the pandemic so far.

% What are the more blue-sky type things? Basically these are things we don't
% really want to do for the first "real" version of sc2ts.
An attractive feature of \texttt{sc2ts} is that the most difficult part of
the inference problem---finding likely recombinant paths through the existing ARG
for new samples---is solved exactly under a well-defined statistical model,
using established HMM methodology (Section~\ref{sec:ls}).
The implementation currently uses
a single, arbitrarily chosen, maximum likelihood path via the
Viterbi algorithm, but there are numerous ways in which the HMM machinery
could be extended in order to explore the set of possible paths, or to
quantify the uncertainty around it. Similarly, the current parameterisation
of the HMM with a single mismatch ratio is simplistic, and it would
be straightforward to condition on per-site mutation rates (and nucleotide-dependent
state transitions, with some additional development).
Recombination breakpoints for the ARG are currently inserted at the
right-most extent of the possible interval
(Section~\ref{sec:breakpoint_intervals}).
More likely locations for the
breakpoint could be chosen within the interval, for example based
on sequence motifs~\citep{Gallaher2020-mr, Yang2020-ct}.
% Hint: there's probably a bunch of other stuff you could do with this
It is likely that the basic machinery of finding matches and quantifying
the uncertainty around them under a well-defined statistical model
in large ARGs would have many applications besides those explored here.

@@

The probabilities of these competing processes of mismatch and recombination are
usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
respectively. For this work we used a slightly different formulation, which
uses one parameter, the mismatch ratio (MMR), to control the relative
importance of mutation and recombination in the HMM. Specifically, an MMR value
of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
 results in copying from a template with no mismatches. To map between
recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths that we wish to be equally likely and
rearrange for the mutation or recombination probability. Consider the simple
case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
consider a region of length $m$. Up until this region, two paths are equally
likely, so we can re-scale by the likelihood of observing the focal sequence up
to the site before the region starts, $c$. For an MMR of $k$, we only need to
consider two of the $n$ members of the reference panel.
One, $\mathcal{P}_1$,
for which there are no mismatches, but we need to recombine in, and a second,
$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
region.
The probability of tracking along each of those paths is given by:
\begin{align*}
\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
\alpha^{m-1}\left(1-\mu\right)^m
&\text{recombine to a template with no mismatches,}\\
\mathbb{P}[\mathcal{P}_2] &= c
\alpha^{m}
% \alpha \alpha^{m-1}
\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
% Note: brackets here emphasise the two different components
% of alpha: (1-r) = don't leave this reference member,
% and r/n = move out, but back to yourself.
where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a
given position with any of the other $n-1$ members of the reference panel
(Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
likelihood of the path up to this point (assumed to be
equal by construction).
We then
set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
Thus, for lower MMR values ($k$ here),
recombination is increasingly favoured over multiple mutations in a specific ancestral genome.
We use an MMR value of $k=3$ in this work,
because of the relatively high rate of recombination relative
to mutation typical of
coronaviruses~\citep{amoutzias2022remarkable}.
Exploring different mismatch ratios and more sophisticated parameterisations
of the HMM are important avenues for future work.

@@

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterisation of using
a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}) and likely
results in a flat likelihood space where many recombinant paths have equal
probability of being chosen.
Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are chosen
from within the possible interval, rather than the current approach of taking
the rightmost value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability to identify recombinants and to filter poor quality data.

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focussing on two likely false positive
recombination nodes at the start of the Delta wave
(``focal'' nodes, in red, corresponding to rows A and B in
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded:
this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, 10\% of Delta
samples are mostly descendants of the node \texttt{tsk232088} on the far right,
a sibling of the earliest focal recombination node.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}

>>>>

\vspace{-1em}
\begin{abstract}
Recombination is an ongoing and increasingly important feature of circulating
lineages of SARS-CoV-2, challenging how we represent the evolutionary history
of this virus and giving rise to new variants of potential public health
concern by combining transmission and immune evasion properties of different
lineages. Detection of new recombinant strains is challenging, with most
methods looking for breaks between sets of mutations that characterise distinct
lineages.
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real time and to cohesively integrate them into a
genealogy in the form of an ancestral recombination graph (ARG),
which jointly records mutation, recombination and genetic
inheritance. We infer two ARGs under
different sampling strategies, and study their properties.
One contains 1.27 million sequences
sampled up to June 30, 2021, and the second is more sparsely sampled,
consisting of 657K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and feature-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
genetic variation, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to utilise an efficient and convenient platform for computation
based on well-engineered technologies
makes \texttt{sc2ts} a promising approach.
\end{abstract}

@@

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from Nextstrain.
We illustrate the backbone phylogeny by visualising
a single tree in the middle of the viral genome,
although other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the Nextstrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

The current version of \texttt{sc2ts} infers a
large number of ``reversion'' mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation (see Section~\ref{sec:args} for precise definitions).
These are symptomatic of both data quality issues
such as ``time travellers'' (Section~\ref{sec:filtering_time_travellers})
and problematic sites (e.g., 28,271 as discussed in
Section~\ref{sec:breakpoint_intervals}; see also
Figure~\ref{fig:pango_XB_gisaid_graph} for an example of multiple
reversions at this site), as well as indicating opportunities
for improvement in tree building heuristics
(Section~\ref{sec:parsimony-heuristics}).
For example, the current ``reversion push'' operation
only eliminates reversions in the newly added portion of the tree
and not reversions around earlier nodes created by this algorithm.

@@

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. Using Nextclade designations, 15 Pango X lineages are
inferred to have multiple
recombinant origins, such that their samples trace to more than one most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e., there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend and which agrees with the official Pango designated parent lineages
(see Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

@@

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps,
is a signature feature of
SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterization of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will be a poor match to
the existing ARG, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
segments.
This appears to be the case for the Delta VoC,
corresponding to rows A and B of Table~\ref{tab:false_positive},
which emerged and quickly rose to high prevalence in early-mid 2021,
carrying 30 characteristic mutations compared to the reference sequence
\citep{McCrone2022-context}, including nine mutations in the S gene
not seen in earlier VoCs.
The origins of the  Delta VoC are
are illustrated in  Figure~\ref{fig:false_positives},
which shows the large numbers of mutations involved,
and the multiple closely related recombinations inferred
before tree-like behaviour is resumed at the ancestor of
[FIXME what percentage?] Delta samples (node \texttt{tsk261771}).
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

@@

% PARA 2. Here are things that could be improved in the short term. Here's a
% few more, longer term ideas.
Nonetheless, \texttt{sc2ts} is currently ``alpha'' quality software, and
we caution against over interpreting current results. As we have sought
to illustrate throughout, there are some clear areas for improvement.
The pipeline used to identify and
mask erroneous sites in the input alignments is simplistic, and, among
other issues, results in multi-base indels being marked as missing data
(Section~\ref{sec:data_preprocessing}). A more sophisticated
approach~\citep[e.g.,][]{Aksamentov2021-hj} would
likely yield significant improvements and reduce the effect of sites with
artefactually high levels of mutation (e.g., site 28271;
Section~\ref{sec:breakpoint_intervals}).
Using a pre-existing tree built using state-of-the-art phylogenetic methods
for the early stages of the pandemic (Section~\ref{sec:backbone_phylogeny}),
and minor adaptations to standard node-dating methods
(Section~\ref{sec:node_dating}) should help resolve the most notable issues with
the inferred backbone phylogeny (Figure~\ref{fig:cophylogeny}).
Trees constructed from daily sample clusters have a surprisingly large
influence on the overall ARG topology
(Section~\ref{sec:sample-cluster-tree-inference}), and so using a more
sophisticated tree building approach should yield clear improvements.
The unrealistically large number of reversion mutations
(Section~\ref{sec:mutation_spectrum}) may be reduced by
improvements to the current parsimony
heuristics (Section~\ref{sec:parsimony-heuristics}).
A major source of errors are the ``time-traveller'' samples, whose
recorded collection dates are months (or years) too early
(Section~\ref{sec:filtering_time_travellers}).
% Hmm, not sure about this. Wouldn't this also rule out new major lineages
% also (which tend to show up with a bunch of new mutations?) I guess it
% depends on the threshold, and we could set it according to previous
% experience with new lineages?
While it is unclear how
we might solve this problem in general, some simple solutions such
as filtering out sequences that exceed a given cost in the LS HMM
(i.e., number of mutations and recombination switches) may work well
in practice. Such an approach would also reduce the impact of sequences
with high levels of sequencing error (which currently contribute a large number of
mutations).
Taken together, these and other relatively minor improvements should
enable inference over much larger subsets of the dataset and
give a clearer picture of the combined processes of recombination and
mutation over the pandemic so far.

% What are the more blue-sky type things? Basically these are things we don't
% really want to do for the first "real" version of sc2ts.
An attractive feature of \texttt{sc2ts} is that the most difficult part of
the inference problem---finding likely recombinant paths through the existing ARG
for new samples---is solved exactly under a well-defined statistical model,
using established HMM methodology (Section~\ref{sec:ls}).
The implementation currently uses
a single, arbitrarily chosen, maximum likelihood path via the
Viterbi algorithm, but there are numerous ways in which the HMM machinery
could be extended in order to explore the set of possible paths, or to
quantify the uncertainty around it. Similarly, the current parameterization
of the HMM with a single mismatch ratio is simplistic, and it would
be straightforward to condition on per-site mutation rates (and nucleotide-dependent
state transitions, with some additional development).
Recombination breakpoints for the ARG are currently inserted at the
right-most extent of the possible interval
(Section~\ref{sec:breakpoint_intervals}).
More likely locations for the
breakpoint could be chosen within the interval, for example based
on sequence motifs~\citep{Gallaher2020-mr, Yang2020-ct}.
% Hint: there's probably a bunch of other stuff you could do with this
It is likely that the basic machinery of finding matches and quantifying
the uncertainty around them under a well-defined statistical model
in large ARGs would have many applications besides those explored here.

@@

The probabilities of these competing processes of mismatch and recombination are
usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
respectively. For this work we used a slightly different formulation, which
uses one parameter, the mismatch ratio (MMR), to control the relative
importance of mutation and recombination in the HMM. Specifically, an MMR value
of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
 results in copying from a template with no mismatches. To map between
recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths that we wish to be equally likely and
rearrange for the mutation or recombination probability. Consider the simple
case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
consider a region of length $m$. Up until this region, two paths are equally
likely, so we can re-scale by the likelihood of observing the focal sequence up
to the site before the region starts, $c$. For an MMR of $k$, we only need to
consider two of the $n$ members of the reference panel.
One, $\mathcal{P}_1$,
for which there are no mismatches, but we need to recombine in, and a second,
$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
region.
The probability of tracking along each of those paths is given by:
\begin{align*}
\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
\alpha^{m-1}\left(1-\mu\right)^m
&\text{recombine to a template with no mismatches,}\\
\mathbb{P}[\mathcal{P}_2] &= c
\alpha^{m}
% \alpha \alpha^{m-1}
\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
% Note: brackets here emphasise the two different components
% of alpha: (1-r) = don't leave this reference member,
% and r/n = move out, but back to yourself.
where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a
given position with any of the other $n-1$ members of the reference panel
(Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
likelihood of the path up to this point (assumed to be
equal by construction).
We then
set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
Thus, for lower MMR values ($k$ here),
recombination is increasingly favoured over multiple mutations in a specific ancestral genome.
We use an MMR value of $k=3$ in this work,
because of the relatively high rate of recombination relative
to mutation typical of
coronaviruses~\citep{amoutzias2022remarkable}.
Exploring different mismatch ratios and more sophisticated parameterizations
of the HMM are important avenues for future work.

@@

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterization of using
a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}) and likely
results in a flat likelihood space where many recombinant paths have equal
probability of being chosen.
Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are chosen
from within the possible interval, rather than the current approach of taking
the rightmost value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability to identify recombinants and to filter poor quality data.

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focusing on two likely false positive
recombination nodes at the start of the Delta wave
(``focal'' nodes, in red, corresponding to rows A and B in
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded:
this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, 10\% of Delta
samples are mostly descendants of the node \texttt{tsk232088} on the far right,
a sibling of the earliest focal recombination node.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}
